Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-24 main.tex
tweaked intro

This is where \emph{combinatorial proofs} come in. They have been
introduced by Hughes~\cite{hughes:pws} for classical propositional
logic as a ``syntax-free'' notion of proof, and as a potential solution
to Hilbert's 24th problem~\cite{hughes:invar} (see
also~\cite{str:hilbert:24}). The basic idea is to abstract away from
the syntax of the inference rules used in the proof and consider the
proof as a combinatorial object, more precisely as a special kind of
graph homomorphism. For example, a propositional combinatorial
proof of Peirce's law $\peirceimpliesformula$ is depicted below-left.
%
\begin{center}\vspace{.5ex}\begin{tikzpicture}\newcommand\radius{2.3}
  \begin{scope}[shift={(-\radius,0)}]{\peircecp}\end{scope}
  \begin{scope}[shift={(\radius,0)}]{\drinkerfibcoloured}\end{scope}
\end{tikzpicture}\vspace{0ex}\end{center}
%
Several authors have illustrated how syntactic proofs in various
proof systems can be translated to propositional combinatorial proofs:
for sequent proofs in~\cite{hughes:invar}, for deep inference proofs
in~\cite{str:fscd17}, for Frege systems in~\cite{str:RR-9048}, and for
tableaux systems and resolution in~\cite{acc:str:18}. This enables a
natural definition of proof identity for propositional logic: two
proofs are \emph{the same}, if they are mapped to the same
combinatorial proof.

@@

In this paper we advance the idea that combinatorial proofs can provide
a notion of proof identity for first-order logic. \emph{First-order
combinatorial proofs} were introduced by Hughes in~\cite{hughes:fopws}.
For example, a first-order combinatorial proof of Smullyan's 
``drinker paradox'' $\drinkerformula$ is shown above-right.
However, even though Hughes proves soundness and completeness, the 
proof is unsatisfactory: (1) the soundness argument is extremely long,
intricate and cumbersome, and (2) the completeness proof does not allow
a syntactic
proof to be read back from a combinatorial proof. In fact, there is the
fundamental problem that not all combinatorial proofs can be obtained
as translations of sequent calculus proofs.

>>>>

This is where \emph{combinatorial proofs} come in. They have been
introduced by Hughes~\cite{hughes:pws} for classical propositional
logic as a ``syntax-free'' notion of proof, and as a potential solution
to Hilbert's 24th problem~\cite{hughes:invar} (see
also~\cite{str:hilbert:24}). The basic idea is to abstract away from
the syntax of the inference rules used in the proof and consider the
proof as a combinatorial object, more precisely as a special kind of
graph homomorphism. For example, a propositional combinatorial
proof of Peirce's law $\peirceimpliesformula$ is depicted below-left,
a homomorphism from a coloured graph to a graph labelled with
propositional variables.
%
\begin{center}\vspace{.5ex}\begin{tikzpicture}\newcommand\radius{2.3}
  \begin{scope}[shift={(-\radius,0)}]{\peircecp}\end{scope}
  \begin{scope}[shift={(\radius,0)}]{\drinkerfibcoloured}\end{scope}
\end{tikzpicture}\vspace{0ex}\end{center}
%
Several authors have illustrated how syntactic proofs in various
proof systems can be translated to propositional combinatorial proofs:
for sequent proofs in~\cite{hughes:invar}, for deep inference proofs
in~\cite{str:fscd17}, for Frege systems in~\cite{str:RR-9048}, and for
tableaux systems and resolution in~\cite{acc:str:18}. This enables a
natural definition of proof identity for propositional logic: two
proofs are \emph{the same}, if they are mapped to the same
combinatorial proof.

@@

In this paper we advance the idea that combinatorial proofs can provide
a notion of proof identity for first-order logic. \emph{First-order
combinatorial proofs} were introduced by Hughes in~\cite{hughes:fopws}.
For example, a first-order combinatorial proof of Smullyan's 
``drinker paradox'' $\drinkerformula$ is shown above-right,
a homomorphism from a partially coloured graph to a labelled graph.
However, even though Hughes proves soundness and completeness, the 
proof is highly unsatisfactory: (1) the soundness argument is extremely
long, intricate and cumbersome, and (2) the completeness proof does not 
allow a syntactic proof to be read back from a combinatorial proof, i.e.,
completeness is not \emph{full}.
A fundamental problem is that not all combinatorial
proofs can be obtained as translations of sequent calculus proofs.
